---
title: "Week 6-Content"
author: "Armand Surbakti"
date: "2025-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## DATA202 - Week 6 Content
### Data Visualization Using `ggplot`

The `ggplot2` package provides alternate functionality for visualizing data. We will examine graphs that have previously been covered in this course but consider how to use functions in the `ggplot2` package to produce them.

```{r}
library(ggplot2)
library(tibble)
```

### Example: 2015 Rugby World Cup Players

To illustrate `ggplot2` functionality, we will use data on players from the 2015 rugby world cup. Let's start by reading and the dataset.
```{r}
rugby <- read.csv("rugby_world_cup_2015.csv")
rugby <- as_tibble(rugby)
names(rugby)
```

```{r}
head(rugby)
```
The dataset contains a lot of information of the rugby metrics.

the team that the player played for (team),
the name of the player (name),
the number of caps for the player (caps),
the player’s position (position),
the time in (fractional) years since the player’s debut (years_since_debut),
the date on which the player debuted (debut),
the age of the player in years (age),
the approximate age of the player when they debuted (approx_age_debuted),
the height of the player in centimeters (height_cm), and
the weight of the player in kilograms (weight_kg).


There are 3 players the name of the team of the team for which 
```{r}
rugby <- rugby %>%
  filter(!(grepl("Utility", position)), team!= " ")
```

Most of the variables in the dataset are numeric, but a few are categorical and saved as character vectors (`team`, `name`, `position`) and team name (team), it makes sense to convert these to factors, as we have unique levels repeated a number of times for each variable. 

When converting the `position` variable to a factor, we will want to specify the ordering of levels to ensure that `R` uses our specified ordering rather than simply ordering these alphanumerically.

```{r}
rugby$position <- factor(rugby$position,
                         levels = c("Hooker", "Prop", "Lock", "Back Row", "Scrum Half", "Fly Half", "Centre", "Wing", "Full Back"))
rugby$team <- factor(rugby$team)
head(rugby)
```
### Visualising a single variable
Before looking at specific graphical displays for numeric and categorical variables 






